---
title: "Stabilizing_selection_analysis"
author: "Siraj"
date: "2025-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading the data I got from Neda from for both the SS and control

```{bash}
cd /Users/selmasri/projects/neda_data/data
ls *.csv
```

````{r}
neda_data_ss <- read.csv("/Users/selmasri/projects/neda_data/data/sieving_count_data.csv", header = TRUE, sep = ";")
str(neda_data_ss)
neda_data_control <- read.csv("/Users/selmasri/projects/neda_data/data/sieving_count_data_controls.csv", 
                              header = TRUE, sep = ";")
str(neda_data_control)
str(neda_data_ss)
````
## function for data cleaning and transformation for both data
```{r}
# Convert columns to formats useful for later:

clean_data <- function(data) {
  data$treatment   <- factor(data$treatment)
  data$replicate   <- factor(data$replicate)
  data$sieve_num   <- factor(data$sieve_num)
  data$sieve_size  <- as.numeric(data$sieve_size)
  data$who         <- factor(data$who)
  data$date        <- as.Date(data$date, format = "%d.%m.%y") 
  data$generation  <- factor(data$generation)
  return(data)
}
neda_data_ss <- clean_data(neda_data_ss)
neda_data_control <- clean_data(neda_data_control)
str(neda_data_control)
```

#Transforming female data into pseudo-continuous by assigning a row with the sieve size value for each fly in a replicate

```{r}
create_pseudo_continuous_female <- function(data) {
  library(dplyr)
  library(tidyr)
  data %>% 
    uncount(female_num, .remove = FALSE) %>%
    mutate(sieve_expanded = sieve_size) %>%
    select(generation, replicate, sieve_expanded, who)
}
pseudo_continuous_female_ss <- create_pseudo_continuous_female(neda_data_ss)
pseudo_continuous_female_control <- create_pseudo_continuous_female(neda_data_control)

```
###Computing pop size for each replicate and including it in the pseudo continuous, in case needed for later
```{r}

pop_size_data_ss <- neda_data_ss %>%
  group_by(generation, replicate) %>%
  summarise(pop_size = sum(female_num, na.rm = TRUE) + sum(male_num, na.rm = TRUE), .groups = "drop")

##including population size in the pseudo continous data for later analysis
pseudo_continuous_female_ss <- pseudo_continuous_female_ss %>%
  left_join(pop_size_data_ss, by = c("generation", "replicate"))

````

####plot the pop size per generation

````{r}
 ggplot(pop_size_data_ss, aes(x = as.factor(generation), y = pop_size, color = replicate, group = replicate)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Population Size by Replicate and Generation",
    x = "Generation",
    y = "Population Size"
  )

```

#Visualisation of replicates across generations in two ways, proportion and pseudo continuous
```{r}
library(Cairo)
library(RColorBrewer)


palette.12 <- brewer.pal(n = 12, name = "Paired")

pseudo_continuous_ss <- ggplot(pseudo_continuous_female_ss, aes(x = factor(generation), y = sieve_expanded, fill = factor(generation))) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(alpha = 0.7, width = 0.1) +
  labs(title = "Violin and Boxplot of Sieve Distributions Across Generations",
       x = "Generation", y = "Sieve Bin (Pseudo-Continuous)", fill = "Generation") +
  scale_fill_manual(values = palette.12) +
  facet_wrap(~ replicate, ncol = 3) +
  theme_minimal()
png(filename = "/Users/selmasri/projects/neda_data/figs/pseudo_continuous_ss.png",
    width = 8, height = 6, units = "in", res = 300, bg = "white")
print(pseudo_continuous_ss)
dev.off()

# ---- Boxplot of Female Proportions Across Generations ----

informative_bins_ss <- neda_data_ss %>%
  group_by(generation, replicate) %>%
  mutate(total_females = sum(female_num, na.rm = TRUE)) %>%  # Compute total females per generation & replicate
  ungroup() %>%
  group_by(generation, replicate, sieve_size) %>%
  summarise(female_prop = sum(female_num, na.rm = TRUE) / unique(total_females), .groups = "drop")



 proportions <- ggplot(informative_bins_ss, aes(x = factor(generation), y = sieve_size, fill = factor(generation))) +
  geom_boxplot(aes(weight = female_prop), alpha = 0.7, outlier.shape = NA) +
  labs(title = "Distribution of Female Proportions Across Generations (Faceted by Replicate)",
       x = "Generation", y = "Sieve Size (µm)", fill = "Generation") +
  scale_fill_manual(values = palette.12) +
  facet_wrap(~ replicate, ncol = 3) +
  scale_y_continuous(breaks = c(937,1085, 1250)) +  # Only key sieve sizes
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        strip.text = element_text(size = 12, face = "bold"))

png(filename = "/Users/selmasri/projects/neda_data/figs/proportions_ss.png",
    width = 8, height = 6, units = "in", res = 300, bg = "white")
print(proportions)
dev.off()


```

# Computing an average fly size per replicate per generation as a summary statistic based on pseudo-continuous transformation of the data


```{r}
###I will change 800 to 850 in the psuedo continuous data so the flies 800 dont drag the mean down although there is rarely females in 800 and then calculate a mean fly size
calc_avg_fly_size <- function(pseudo_data) {
  pseudo_data %>%
    mutate(sieve_expanded = ifelse(sieve_expanded == 800, 850, sieve_expanded)) %>%
    group_by(generation, replicate) %>%
    summarise(avg_fly_size = mean(sieve_expanded, na.rm = TRUE), .groups = "drop")
}

##Apply for SS
avg_fly_size_ss <- calc_avg_fly_size(pseudo_continuous_female_ss)
head(avg_fly_size_ss)

#Apply for control
avg_fly_size_control <- calc_avg_fly_size(pseudo_continuous_female_control)
head(avg_fly_size_control)





pseudo_continuous_female_ss_850 <- pseudo_continuous_female_ss %>%
  mutate(sieve_expanded = ifelse(sieve_expanded == 800, 850, sieve_expanded))

avg_fly_size_ss <- pseudo_continuous_female_ss_850 %>%
  group_by(generation, replicate) %>%
  summarise(avg_fly_size = mean(sieve_expanded, na.rm = TRUE), .groups = "drop")  # Compute mean sieve size

head(avg_fly_size_ss)

```

##Visualisation 
```{r}
avg_fly_plot_ss <- ggplot(avg_fly_size_ss, aes(x = factor(generation), y = avg_fly_size)) +
  geom_boxplot(fill = "orange", alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, color = "black") +
  labs(title = "Average Fly Size per Generation", 
       x = "Generation", 
       y = "Mean Fly Size (µm)") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),        # Bold, larger title
    axis.title.x = element_text(size = 14, face = "bold"),        # Bold, larger x-axis title
    axis.title.y = element_text(size = 14, face = "bold"),        # Bold, larger y-axis title
    legend.title = element_text(size = 12, face = "bold"),        # Bold legend title
    legend.text = element_text(size = 10)                         # Larger legend text
  )

ggsave(
  filename = "/Users/selmasri/projects/neda_data/figs/mean_size_ss.png",
  plot = avg_fly_plot_ss,
  width = 8,
  height = 6,
  dpi = 300
)


avg_fly_ss_labeled <- ggplot(avg_fly_size_ss, aes(x = factor(generation), y = avg_fly_size)) +
  geom_boxplot(fill = "orange", alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, aes(color = replicate)) +
  geom_text(aes(label = replicate), 
            position = position_jitter(width = 0.2, height = 0), 
            vjust = -1.5, size = 3.5) +
  labs(title = "Average Fly Size per Generation",
       x = "Generation", y = "Mean Fly Size (µm)") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10)
  )

ggsave(
  filename = "/Users/selmasri/projects/neda_data/figs/mean_size_ss_labels.png",
  plot = avg_fly_ss_labeled,
  width = 8,
  height = 6,
  dpi = 300
)
```

##Control visualisation

```{r}
ggplot(avg_fly_size_control, aes(x = factor(generation), y = avg_fly_size)) +
  geom_boxplot(fill = "green", alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, aes(color = replicate)) +
  geom_text(aes(label = replicate), 
            position = position_jitter(width = 0.2, height = 0), 
            vjust = -1.5, size = 3.5) +
  labs(title = "Average Fly Size per Generation (Increasing Selection Replicates 90->82%)",
       x = "Generation", y = "Mean Fly Size (µm)") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10)
  )


```
#Anova with generation and replicate as predictor and mean fly size as response
```{r}
#the summarize function above produced a tibble and the anova needs a df

avg_fly_size_ss <- as.data.frame(avg_fly_size_ss)

anova_result_ss <- aov(avg_fly_size ~ generation + replicate, data = avg_fly_size_ss)
summary(anova_result_ss)


shapiro.test(residuals(anova_result_ss))  # Should be > 0.05 for normality, which is the case

qqnorm(residuals(anova_result_ss))
qqline(residuals(anova_result_ss), col = "red")
residuals_ss <- residuals(anova_result_ss)

library(car)
leveneTest(residuals_ss ~ avg_fly_size_ss$generation) #p> 0.05 assumption met of homogenity of variance.
TukeyHSD(anova_result_ss)

```
