---
title: "Measurements_analysis"
author: "Siraj"
date: "2025-03-17"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/selmasri/projects/trunc_sel_ee/data/phenotypic/raw")
```
# Phenotypic data preparation

We start with our data stored in the following csv files:


```{bash}
cd /Users/selmasri/projects/trunc_sel_ee/data/phenotypic/raw
ls *.csv

```
Setting working directory 

```{r}
setwd("/Users/selmasri/projects/trunc_sel_ee/data/phenotypic/raw")

```
Next, I process the data, having and I calculate the mean for the the several measurments of the trait by summarizing the several rows corresponding to the fly id and trait

# loading the data
```{r}
library(emmeans)
library(factoextra)
library(corrplot)
library(psych)
library(dplyr)
library(ggplot2)
library(car)
library(tidyr)
library(dplyr)
library(GGally)
library(lmtest)

main_df <- read.csv("Thorax_measurments.csv", header =  TRUE, sep = ",")
str(main_df)
main_df$Date_measured <- as.Date(main_df$Date_measured, format = "%d/%m/%Y")
str(main_df)
# Aggregating replicate measurements by Fly_ID (and Date_measured if needed)
main_df_tidy <- main_df %>%
  group_by(Fly_ID, Date_measured) %>%
  summarise(
    Thorax_length_avg = mean(Thorax_length.mm., na.rm = TRUE),
    Full_body_length_avg = mean(Full_body_length.mm., na.rm = TRUE),
    Thorax_width_avg = mean(Thorax_width.mm., na.rm = TRUE),
    Wing_length_avg = mean(wing.length.mm., na.rm = TRUE)
  )

```
Now there is another part of the data 28 flies present in other files each fly has a .csv file(because of semi automation) i had to check all these individual measurments (11 per fly) to see if they correspond to the right trait since they are not labeled, now the code will average the first 3 rows as thorax length then second 3 as body length etc 

##loading the 2nd set of data
```{r}
# Path to the folder of individual CSVs
path <- "/Users/selmasri/Desktop/Project/measurments_results_gwas1"

# List all CSV files in that directory
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)

# A small helper function to parse Fly_ID from a filename like "fly_0189_results.csv"
get_fly_id <- function(filename) {
  base <- basename(filename)               # e.g. "fly_0189_results.csv"
  no_ext <- sub("\\.csv$", "", base)       # "fly_0189_results"
  fly_id <- sub("_results$", "", no_ext)   # "fly_0189"
  return(fly_id)
}

# A helper function that safely averages the "Length" column for a given set of row indices.
# If the file doesn't have enough rows, it returns NA.
get_mean <- function(df, idx, col = "Length") {
  if (all(idx <= nrow(df))) {
    mean(df[idx, col], na.rm = TRUE)
  } else {
    NA
  }
}

list_of_dfs <- lapply(files, function(file) {
  # 1) Parse the Fly_ID from the filename
  fly_id <- get_fly_id(file)
  
  # 2) Read the CSV
  df <- read.csv(file)
  
  # 3) For each trait, average the correct rows 
  thorax_length_avg   <- get_mean(df, 1:3,  "Length")
  body_length_avg     <- get_mean(df, 4:6,  "Length")
  thorax_width_avg    <- get_mean(df, 7:9,  "Length")
  wing_length_avg     <- get_mean(df, 10:11,"Length")  # may be NA if only 9 rows
  
  # 4) Create a single-row data frame with all traits
  out <- data.frame(
    Fly_ID             = fly_id,
    Date_measured      = as.Date("2025-01-27"),  # Hard-code this date
    Thorax_length_avg  = thorax_length_avg,
    Full_body_length_avg = body_length_avg,
    Thorax_width_avg   = thorax_width_avg,
    Wing_length_avg    = wing_length_avg
  )
  
  return(out)
})

# Combine all mini data frames for flies 188–215 into one
other_df <- bind_rows(list_of_dfs)

# Check the result
str(other_df)
head(other_df)




```

##Merging the two dataframes

```{r}
# 1) Remove flies 188 through 215 from the main data frame(note that original one had empty rows from 188 to 215)
#    Here, we're filtering out any Fly_ID that matches "fly_0188" through "fly_0215".
main_df_no_placeholders <- main_df_tidy %>%
  filter(
    !(Fly_ID %in% paste0("fly_0", 188:215))
  )

# 2) Stack the main data (with placeholders removed) and the other data
all_flies_df <- bind_rows(main_df_no_placeholders, other_df)

# 3) Inspect the final combined data
str(all_flies_df)
head(all_flies_df)


```
#Correlation between measurments
```{r}
# ---- Correlation between measurments ----
##correlation between measurments
#creating a vector with column of interest
measure_cols <- c("Thorax_length_avg", "Full_body_length_avg", 
                  "Thorax_width_avg", "Wing_length_avg")
#subseting for columns of interest
sub_df <- all_flies_df[, measure_cols]

# Perform pairwise correlation with corr.test() storing in res_psych to use,
#the psych pacakge beacuse many rows have only two obersvations and some have 3 some have 4, and i dont want to omit any data
#bonferroni p value correction
res_psych <- corr.test(sub_df, use = "pairwise", method = "pearson", adjust = "bonferroni")

# View the results
res_psych$r   # Correlation matrix
res_psych$n   # Number of observations used for each pair
res_psych$p   # p-values for each correlation
summary(sub_df)
range(sub_df$Thorax_length_avg, na.rm = TRUE)
range(sub_df$Thorax_width_avg, na.rm = TRUE)
range(sub_df$Wing_length_avg, na.rm = TRUE)
```




###Visualizing using pairsplot
```{r}
pairsplot <- ggpairs(sub_df,
                     # Lower panels: scatterplots with a linear smoothing line and transparent points
                     lower = list(continuous = wrap("smooth", method = "lm", se = FALSE, na.rm = TRUE, alpha = 0.4)),
                     # Diagonal panels: density plots or histograms
                     diag = list(continuous = "densityDiag"),
                     # Upper panels: display correlation coefficients using pairwise complete observations
                     upper = list(continuous = wrap("cor", use = "pairwise.complete.obs", method= "pearson")),
                     title = "Pairs Plot of Fly Traits"
) + theme_bw()

# Display the plot
pairsplot

# Save the plot to a file
ggsave(
  filename = "/Users/selmasri/projects/trunc_sel_ee/figs/pairsplotphenotypes.png",
  plot = pairsplot,
  width = 10,
  height = 10,
  units = "in"
)


```

##loading full fly information
First I have to load a mapping file which is file mapping the anonymous fly ID to the original ID which contain the bin information. Then mearge it with the full data frame to add the fly info
```{r}
# load the mapping file 
mapping <- read.csv("/Users/selmasri/Desktop/Project/mapping.csv", stringsAsFactors = FALSE)

# Check the first rows
head(mapping)

### I will merge the data so i know later which fly has which id

# adding .tif extension to Fly_ID so the merging works
all_flies_df$Fly_ID <- ifelse(grepl("\\.tif$", all_flies_df$Fly_ID),
                              all_flies_df$Fly_ID,
                              paste0(all_flies_df$Fly_ID, ".tif"))

# merge the data 
merged_data <- merge(all_flies_df, mapping, 
                     by.x = "Fly_ID", by.y = "Anonymized_ID", 
                     all.x = TRUE)

# Check the merged data again
head(merged_data)
# Remove the ".tif" extension from Original_File
merged_data$original_id_clean <- gsub("\\.tif$", "", merged_data$Original_File)

# Split the cleaned original_id into parts using the underscore "_" as the delimiter
meta_list <- strsplit(merged_data$original_id_clean, "_")

# Convert the list to a data frame; we expect 5 parts: Batch, SieveSize, Plate, Column, and Row.
meta_df <- do.call(rbind, meta_list)
colnames(meta_df) <- c("batch", "sieve_size", "plate", "column", "row")

# Convert meta_df to a proper data frame
meta_df <- as.data.frame(meta_df, stringsAsFactors = FALSE)

# Combine the metadata with merged_data
merged_data <- cbind(merged_data, meta_df)

# Check the merged data to see the new metadata columns
head(merged_data)


```
##transforming the data into a long format by trait
This will help us to omit the NAs without omitting a whole fly

```{r}
long_data <- merged_data %>%
  pivot_longer(cols = c("Thorax_length_avg", "Full_body_length_avg", 
                        "Thorax_width_avg", "Wing_length_avg"),
               names_to = "Trait",
               values_to = "Value")



long_data$sieve_size <- factor(long_data$sieve_size, levels = sort(unique(long_data$sieve_size)))

```
##Visualisation of the traits distribution and their distribution by sieve
```{r}

ggplot(long_data, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~ Trait, scales = "free") +
  labs(
    title = "Distribution of Fly Traits",
    x = "Measurement Value (mm)",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(face = "bold", size = 14),
    axis.text.x  = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 14),
    axis.text.y  = element_text(face = "bold", size = 12),
    plot.title   = element_text(face = "bold", size = 16, hjust = 0.5)
  )

traits_dis_by_sieve_size <- ggplot(long_data, aes(x = factor(sieve_size, levels = sort(unique(as.numeric(as.character(sieve_size))))), 
                                                  y = Value, fill = factor(sieve_size))) +
  geom_boxplot() +
  facet_wrap(~ Trait, scales = "free_y") +
  labs(x = "Sieve Size", y = "Measurement",
       title = "Fly Traits by Sieve Size") +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")  # Make facet labels bold & bigger
  )

ggsave(
  filename = "/Users/selmasri/projects/trunc_sel_ee/figs/traits_by_sieve_size.png",
  plot = traits_dis_by_sieve_size,
  width = 15,
  height = 10,
  units = "in"
)

```
###PCA
The PCA loading full body and wing length show almost a corr of 1 while in the pairs plot its around 0.5?
```{r}
##omit NAs for the pca analysis to have complete data for the 4 traits
pca_df <- na.omit(sub_df)
pca_res <- prcomp(pca_df, center = TRUE, scale. = TRUE)
summary(pca_res)
pca_res$rotation #gives the loadings (i.e., how each original variable contributes to each PC).
pca_res$x #gives the coordinates of each observation (fly) in the new PCA space.

# Calculate percentage of variance for PC1 and PC2 then plot
eig.val <- pca_res$sdev^2
var.percent <- eig.val / sum(eig.val) * 100
pc1 <- round(var.percent[1], 1)
pc2 <- round(var.percent[2], 1)

 fviz_pca_biplot(pca_res,
                label = "var",
                repel = TRUE,
                geom = "point") +
  labs(
    x = paste0("PC1 (", pc1, "%)"),
    y = paste0("PC2 (", pc2, "%)")
  ) +
  theme_classic() +                                # Use a classic theme for clearer axes
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +  # Add horizontal axis line
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey")      # Add vertical axis line
 
 
# I can color the scale by sieve size, maybe also a pallete of a spectrum to see if there is a gradient
 
 #fviz_pca_biplot(pca_res,
#label = "var",      # label only variables (arrows)
                #repel = TRUE,       # reduce label overlap
                #geom = "point",     # plot individuals as points
                #habillage = grouping,   # color points by sieve size
                #addEllipses = TRUE,      # optional: add confidence ellipses
                #palette = "jco") +       # or use any palette you prefer
  # labs(x = paste0("PC1 (", pc1, "%)"),
#y = paste0("PC2 (", pc2, "%)"),
#title = "PCA Biplot Colored by Sieve Size") +
  # theme_classic() +
  #geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  # geom_vline(xintercept = 0, linetype = "dashed", color = "grey")





```

#Linear Models

```{r}
# 1. Ensure sieve_size is a factor (adjust levels if needed)
# convert to a factor
merged_data$sieve_size <- factor(merged_data$sieve_size, 
                                 levels = sort(unique(merged_data$sieve_size)))
 levels(merged_data$sieve_size)
 
 # Replace the level "800" with "850" so the value doesnt drive significance
 levels(merged_data$sieve_size)[levels(merged_data$sieve_size) == "800"] <- "850"
 
 
 levels(merged_data$sieve_size)
 # (Optional) Reorder levels if needed
 merged_data$sieve_size <- factor(merged_data$sieve_size, levels = sort(unique(merged_data$sieve_size)))
 
##the last sieve was showing a decrease for most flies due to techincal issue because we notice it has little flies and mostly males so rerun the association by removing extremes
# Remove rows where sieve_size is 800, 1319, or 1400
merged_data_noextremes <- merged_data[!(merged_data$sieve_size %in% c(800, 1319, 1400)), ]

# Ensure sieve_size is still treated as a factor with updated levels
merged_data_noextremes$sieve_size <- factor(merged_data_noextremes$sieve_size)

# Check the unique sieve sizes remaining
unique(merged_data_noextremes$sieve_size)
# Fit models for each trait
lm_thorax <- lm(Thorax_length_avg ~ sieve_size, data = merged_data)
lm_body   <- lm(Full_body_length_avg ~ sieve_size, data = merged_data)
lm_width  <- lm(Thorax_width_avg ~ sieve_size, data = merged_data)
lm_wing <- lm(Wing_length_avg ~ sieve_size, data = merged_data)
lm_wing_noextremes   <- lm(Wing_length_avg ~ sieve_size, data = merged_data_noextremes)
lm_thorax_date <- lm(Thorax_length_avg ~ sieve_size + as.numeric(Date_measured), data = merged_data)
summary(lm_thorax_date)
anova(lm_thorax, lm_thorax_date)
summary(lm_thorax)


merged_data$DateFactor <- factor(merged_data$Date_measured)
lm_thorax_date_factor <- lm(Thorax_length_avg ~ sieve_size + DateFactor, data = merged_data)
summary(lm_thorax_date_factor)
table_counts <- table(merged_data$Date_measured, merged_data$sieve_size)

print(table_counts)

chisq_result <- chisq.test(table_counts)

print(chisq_result)  ####conclusion the high inflation in R2 result with date measured included is a bias caused by dependance(OR CORR) between sive ize and date measured in this data

# Extract adjusted R-squared values for each model.
adj_r2_thorax <- summary(lm_thorax)$adj.r.squared
adj_r2_body   <- summary(lm_body)$adj.r.squared
adj_r2_width  <- summary(lm_width)$adj.r.squared
adj_r2_wing   <- summary(lm_wing)$adj.r.squared

# Print the values
print(paste("Adjusted R² for Thorax:", adj_r2_thorax))
print(paste("Adjusted R² for Body:", adj_r2_body))
print(paste("Adjusted R² for Width:", adj_r2_width))
print(paste("Adjusted R² for Wing:", adj_r2_wing))

adj_r2_values <- data.frame(
  Trait = c("Thorax", "Body", "Width", "Wing"),
  Adj_R2 = c(adj_r2_thorax, adj_r2_body, adj_r2_width, adj_r2_wing)
)

# Fit linear models with filtered data
lm_thorax_filtered <- lm(Thorax_length_avg ~ sieve_size, data = merged_data_noextremes)
lm_body_filtered  <- lm(Full_body_length_avg ~ sieve_size, data = merged_data_noextremes)
lm_width_filtered  <- lm(Thorax_width_avg ~ sieve_size, data = merged_data_noextremes)
lm_wing_filtered   <- lm(Wing_length_avg ~ sieve_size, data = merged_data_noextremes)

# Model summaries
summary(lm_thorax_filtered)
summary(lm_body_filtered)
summary(lm_width_filtered)
summary(lm_wing_filtered)
#####COCNLUSION FILTERING THE DATA DOES NOT CHANGE THE MOST ASSOCIATED TRAIT SO IT IS NOT USEFUL THE CODE ABOVE OF FILTERING ETC

```

##Model diagnostics

```{r}

# 3. Model diagnostics
# Plot diagnostic plots to check for normality, homoscedasticity, and influential points.
plot(lm_thorax, which = 1)  # Residuals vs. Fitted
plot(lm_thorax, which = 2)
plot(lm_wing, which = 1)
plot(lm_wing, which = 2)
# Normal Q-Q plot
merged_data[28, ]
##fucntion to test model assumptions
test_model_assumptions <- function(model) {
  # 1. Normality of residuals: Shapiro–Wilk
  shapiro_res <- shapiro.test(residuals(model))
  
  # 2. Homoscedasticity: Breusch–Pagan
  bptest_res <- bptest(model)
  
  cat("\nShapiro–Wilk Test for Normality:\n")
  print(shapiro_res)
  
  cat("\nBreusch–Pagan Test for Homoscedasticity:\n")
  print(bptest_res)
}

# Usage example (for your four models):
test_model_assumptions(lm_thorax)
test_model_assumptions(lm_body)
test_model_assumptions(lm_width)
test_model_assumptions(lm_wing)
test_model_assumptions(lm_wing_noextremes)



```

###Log transforming the data to see if there is a change in the model assumptions
```{r}
merged_data$Wing_length_log <- log(merged_data$Wing_length_avg)
merged_data$Thorax_length_log <- log(merged_data$Thorax_length_avg)
lm_wing_log <- lm(Wing_length_log ~ sieve_size, data = merged_data)
lm_thorax_log <- lm(Thorax_length_log ~ sieve_size, data = merged_data)

test_model_assumptions(lm_wing_log) ##did not improve anything
test_model_assumptions(lm_thorax_log)## the residuals are normally distriubted now

adj_r2_thorax_log <- summary(lm_thorax_log)$adj.r.squared
adj_r2_thorax_log ## now the adjusted r squared is a little bit higher than before but stilll lower than wing etc


```


# Function to compare null vs full model for a given trait

```{r}

#I should double check how this function is working
run_model_comparison <- function(trait, data) {
  # Build formulas for the null model and full model
  full_formula <- as.formula(paste(trait, "~ sieve_size"))
  null_formula <- as.formula(paste(trait, "~ 1"))
  
  # Fit the models
  model_null <- lm(null_formula, data = data)
  model_full <- lm(full_formula, data = data)
  
  # Compare the models using ANOVA
  comparison <- anova(model_null, model_full)
  return(comparison)
}

```

###emmeans
```{r}
emm_thorax <- emmeans(lm_thorax, ~ sieve_size)
plot(emm_thorax)

```
###Sort and Re-factor the sieve_size Column
```{r}
# 1) Extract the unique sieve sizes as character
# 2) Convert them to numeric
# 3) Sort them
sorted_levels <- sort(as.numeric(unique(as.character(merged_data$sieve_size))))

# 4) Re-factor sieve_size with these sorted levels
merged_data$sieve_size <- factor(
  merged_data$sieve_size, 
  levels = sorted_levels
)

# Check that the levels are in ascending order
levels(merged_data$sieve_size)


```

####Subset the Data to Rows with Non-missing Wing_length_avg and Drop Unused Factor Levels
```{r}
wing_data <- subset(merged_data, !is.na(Wing_length_avg))
wing_data$sieve_size <- droplevels(wing_data$sieve_size)
levels(wing_data$sieve_size)
```

##Plot Wing Length by Sieve Size
```{r}
ggplot(wing_data, aes(x = sieve_size, y = Wing_length_avg, fill = factor(sieve_size))) +
  geom_boxplot() +
  labs(title = "Wing Length by Sieve Size",
       x = "Sieve Size",
       y = "Wing Length (mm)") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(wing_data, aes(x = factor(sieve_size), y = Wing_length_avg, fill = factor(sieve_size))) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  labs(title = "Wing Length by Sieve Size", 
       x = "Sieve Size", 
       y = "Wing Length (mm)") +
  theme_bw() +
  theme(legend.position = "none")

```

####count the 1250 flies


```{r}
sum(merged_data$sieve_size == "1250")
sum(merged_data$sieve_size == "1250" & !is.na(merged_data$Wing_length_avg))
```